---
title: "my-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

---
title: "TaxoTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TaxoTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Welcome to *TaxoTools*, an R package designed to integrate character, geographic, and phylogenetic data in taxonomy. This vignette provides an overview of the package's key features and demonstrates its usage with example datasets.

## Installation

To install *TaxoTools*, use the following commands:

#### Install from CRAN (if applicable)

install.packages("TaxoTools")

#### Or install the development version from GitHub

devtools::install_github("mschools287/TaxoTools")


## Key Features

TaxoTools includes several key features designed to facilitate taxonomic analysis:

- **Character Data Analysis and Visualization**: Functions for comparing and visualizing traits.
- **Geographic Analysis and Visualization**: Functions for comparing and visualizing locality data.
- **Phylogenetic Tree Analysis and Visualization**: Tools to visualize phylogenetic trees with annotated node information.
- **Taxonomic Monograph Generation**: Functions for generating taxonomic descriptions.

Each of these features is designed to work together to provide a comprehensive toolkit for taxonomists.

## Example Workflow

### 1. Load the Package

```{r}
library(TaxoTools)
library(ape)
library(phytools)
```
### 2. Load example data and root the trees
#### Load the *Sauresia* time tree
```{r}
sauresia_Time <- read.tree("C:\\Users\\mschools\\Documents\\TaxoTools\\vignettes\\examples\\sauresia_48tx_timetree.nwk")
plot(sauresia_Time)
```

#### Load, root, and check the ML *Sauresia* tree
```{r}
sauresia_ML <- read.tree("C:\\Users\\mschools\\Documents\\TaxoTools\\vignettes\\examples\\sauresia_48tx_IQtree_rooted.nwk")
sauresia_ML_rooted <- reroot(
    sauresia_ML, 
    48, 
    position = sauresia_ML$edge.length[which(sauresia_ML$edge[,2]==48)]/2)
is.rooted(sauresia_ML_rooted)
plot(sauresia_ML_rooted, show.node.label = T)
```

#### Load, root, and check the Bayesian *Sauresia* tree
```{r}
sauresia_B <- read.tree("C:\\Users\\mschools\\Documents\\TaxoTools\\vignettes\\examples\\sauresia_48tx_MrBayes_rooted.nwk")
sauresia_B_rooted <- reroot(
    sauresia_B, 
    48, 
    position = sauresia_B$edge.length[which(sauresia_B$edge[,2]==48)]/2)
is.rooted(sauresia_B_rooted)
plot(sauresia_B_rooted, show.node.label = T)
```

#### Load in *Sauresia* morphodata and metadata (unclean and clean version)
##### Uncleaned version
```{r}
sauresia_morpho <- read.csv("C:\\Users\\mschools\\Documents\\TaxoTools\\vignettes\\examples\\sauresia_morpho.csv")
```

##### Cleaned version
```{r}
sauresia_morpho2 <- read.csv("C:\\Users\\mschools\\Documents\\TaxoTools\\vignettes\\examples\\sauresia_morpho2.csv")
```

#### Load in name file with cleaned specimen names
```{r}
new_names <- read.csv("C:\\Users\\mschools\\Documents\\TaxoTools\\vignettes\\examples\\new_tree_names.csv")
```

### 3. Data Cleaning and Processing Functions
#### `process_check_data()`: Check for common errors in the character dataset
```{r}
process_check_data(
  data = sauresia_morpho,
  lat_col = "Latitude..decimal.",
  long_col = "Longitude..decimal.",
  elevation_col = "Elevation",
  trait_columns = c(
    "SVL", "pattern", "head.markings", "nuchal.lines",
    "dots.in.lateral.band", "mental.to.vent.scale.count",
    "midbody.scalecount", "total.finger.lamella", "strigae.on.ten.scales",
    "total.toe.length", "scale.under.eye.to.lip.distance", "eye.length", "arm.length",
    "ear.width", "rostral.width", "rostral.height", "head.length", "mental.width",
    "post.mental.width", "cloaca.width", "prefrontal.width",
    "largest.supraoccular.width", "Finger.III.length", "ear.to.eye.distance",
    "head.width", "frontal.width", "frontal.lenght"),
  date_col = "Date.Collected")
```
#### `process_column_names()`: Change column names to clean versions
```{r}
sauresia_morpho3 <- process_column_names(
    sauresia_morpho2, 
    new_column_names = c(
        "Name in Tree", "New names", "Type", "Genus", "Species", "Museum", "Number",
        "Country", "Province", "Locality", "Elevation",
        "Date Collected", "Collectors", "Remarks", "Latitude", "Longitude",
        "SVL", "Pattern", "Head Markings", "Nuchal Lines", "Dots in Lateral Band",
        "Mental to Vent Scale Count", "Midbody ScaleCount", "Total Finger Lamella",
        "Strigae on Ten Scales", "Total Toe Length", "Scale Under Eye to Lip Distance",
        "Eye Length", "Arm Length", "Ear Width", "Rostral Width", "Rostral Height",
        "Head Length", "Mental Width", "Post Mental Width", "Cloaca Width",
        "Prefrontal Width", "Largest Supraoccular Width", "Finger III Length",
        "Ear to Eye Distance", "Head Width", "Frontal Width", "Frontal Length"))
```
#### `process_juveniles()`: Remove juviniles from the dataset based on their trait size compared to the largest individual in their respective group
```{r}
process_juveniles(
  sauresia_morpho3,
  trait = "SVL",
  group_by = "Species",
  cutoff = 0.85)
```
#### `process_check_missing()`: Check for the number and proportion of missing values in each column
```{r}
process_check_missing(data = sauresia_morpho3)
```
#### `process_convert_units()`: Convert the units of specified traits
```{r}
process_convert_units(
  data = sauresia_morpho3,
  trait_column = "SVL",
  conversion_factor = 10)
```
#### `process_standardize_traits()`: Standardize traits by a reference trait
```{r}
process_standardize_traits(
  data = sauresia_morpho3,
  reference_trait = "SVL",
  traits_to_standardize = c(
    "SVL", 
    "Pattern", 
    "Head Markings",
    "Head Width", "Frontal Width"))
```
#### `process_phylo_count()`: Count the number of individuals per group in a phylogeny
```{r}
process_phylo_count(
  sauresia_ML_rooted,
  group_name = "sepsoides")
```
#### `process_specimen_count()`: Count the number of individualsper group in a character dataset
```{r}
process_specimen_count(
  sauresia_morpho3,
  grouped_by = "Species")
```
#### `process_phylo_names()`: Update individual names in the phylogenies to be more legible and match those in other datasets
```{r}
sauresia_ML_rooted_named <- process_phylo_names(
  sauresia_ML_rooted,
  new_names)
sauresia_B_rooted_named <- process_phylo_names(
  sauresia_B_rooted,
  new_names)
sauresia_T_named <- process_phylo_names(
  sauresia_Time,
  new_names)
```
#### `process_data_matches()`: Check for consistency in individual names between datasets
```{r}
process_data_matches(
  phy1 = sauresia_ML_rooted,
  data1 = sauresia_morpho3,
  data1_column = "Name in Tree")
```
#### process_subset_group(): Subset the character data
```{r}
process_subset_group(
  sauresia_morpho3,
  grouped_by = "Species",
  prefix = "Subset_")
```
### 4. Character (Morphological) Functions
#### `trait_filter()`: Check which individuals fall above/below a threshold for a given trait
```{r}
trait_filter(
  sauresia_morpho3,
  individuals_column = "Number",
  trait_column = "SVL",
  threshold = 60,
  above = TRUE)
```
#### `trait_min_max()`: Check which group has the min and max value for a given trait
```{r}
trait_min_max(
  sauresia_morpho3,
  grouped_by = "Species",
  trait_column = "SVL")
```
#### `trait_summary()`: Compute the summary statistics for a given trait per group
```{r}
trait_summary(
  data = sauresia_morpho3,
  grouped_by = "Species",
  trait_column = "SVL")
```
#### `trait_compare()`: Compare the distribution of a trait between two groups (scatterplots and interactive box plots)
```{r}
trait_compare(
  data = sauresia_morpho3,
  group_by_column = "Species",
  group_values = c("Sauresia cayemitae", "Sauresia pangnolae"),
  traits = c("SVL", "Total Toe Length"))
```
#### `trait_not_separated()`: Check which groups cannot be distinguished pairwise based on the included traits
```{r}
trait_not_separated(
  data = sauresia_morpho3,
  trait_columns = c("SVL", "Pattern", "Head Markings",
                    "Head Width", "Frontal Width"),
  grouped_by = "Species")
```
### 5. Geographic Functions
#### `geo_range()`: Print the max and min longitude and latitude coordinated for each group
```{r}
geo_range(
  sauresia_morpho3,
  grouped_by = "Species",
  lat_col = "Latitude",
  long_col = "Longitude")
```
#### `geo_convex_hull()`: Calculate the area of the convex hull for each group with more than three sets of coordinates
```{r}
geo_convex_hull(
  sauresia_morpho3,
  grouped_by = "Species",
  lat_col = "Latitude",
  long_col = "Longitude",
  area_unit = "km^2")
```
#### `geo_plot_map()`: Plot an interactive map showing occurance data and convex hulls for groups where applicable
```{r}
geo_plot_map(
  sauresia_morpho3,
  grouped_by = "Species",
  lat_col = "Latitude",
  long_col = "Longitude")
```
#### `geo_range_overlaps()`: Print a table showing which groups have overlapping ranges (
###### Calculated using convex hulls (3+ sets of coordinates), lines (2 sets of coordinates), and points (1 set of coordinates)
```{r}
geo_range_overlaps(
  sauresia_morpho3,
  group_by = "Species",
  lat_col = "Latitude",
  long_col = "Longitude")
```
#### `geo_relatedness()`: Check for spatial patterns in phylogenetic relatedness
```{r}
geo_relatedness(
  data = sauresia_morpho3,
  lat_column = "Latitude",
  long_column = "Longitude",
  phylogeny = sauresia_ML_rooted_named,
  reference_species = "Wetmorena surda 1",
  individuals_column = "New names")
```
### 6. Phylogenetic Functions
#### `phylo_tree_summary()`: Get summary stats for the support or time values in a phylogeny
```{r}
phylo_tree_summary(
  tree = sauresia_ML_rooted_named,
  method = "ML")
```
#### `phylo_clade_support()`: Get summary stats for the support or time for a specific group in the phylogeny
```{r}
phylo_clade_support(
  sauresia_ML_rooted_named,
  "sepsoides",
  method = "ML")
```
#### `phylo_divergence_histogram()`: Plot a histogram of divergence dates from a timetree
```{r}
phylo_divergence_histogram(
  sauresia_T_named,
  bin_width = .1,
  color = "lightblue")
```
#### `phylo_taxa_divergence_hist()`: Plot a histogram comparing divergence times between multiple taxa in a timetree
```{r}
phylo_taxa_divergence_hist(
  tree = sauresia_T_named,
  taxa_list = c("sepsoides", "agramma", "cayemitae"),
  colors = c("darkseagreen", "deeppink", "pink"),
  bin_width = 2)
```
#### `phylo_node_support_hist()`: Plot a histogram of node support values from two phylogenies
```{r}
phylo_node_support_hist(
  tree_ml = sauresia_ML_rooted_named,
  tree_bayesian = sauresia_B_rooted_named,
  bin_width = 5,
  colors = c("skyblue", "green"))
```
#### `phylo_heatmap()`: Plot a heatmap of phylogenetic relatedness for all individuals in a phylogeny
```{r}
phylo_heatmap(phylogeny = sauresia_ML_rooted_named)
```
#### `phylo_highlight_taxa()`: Plot multiple phylogenies side-by-side highlighting important taxa and nodes
```{r}
phylo_highlight_taxa(
  trees = list(sauresia_ML_rooted_named, sauresia_B_rooted_named),
  taxa_to_highlight = c("sepsoides", "Grande Cayemite"),
  highlight_colors = c("green", "orange"),
  node_support_threshold = 95,
  support_color = "blue",
  highlight_shape = "circle",
  shape_size = 1,
  tip_label_size = 0.8)
```
#### `phylo_plot_trees()`: Plot multiple phylogenies side-by-side with customization options including the ability to color support values based on significance
```{r}
phylo_plot_trees(
  trees = list(sauresia_ML_rooted_named, sauresia_B_rooted_named, sauresia_T_named),
  methods = c("ML", "Bayesian", "Time Tree"),
  show_bootstraps = T,
  bootstrap_color = "colored",
  show_posterior = T,
  posterior_color = "colored",
  time_scale = T,
  side_by_side = T,
  name_data = new_names,
  phylogeny_type = "phylogram",
  edge_color = "blue")
```
#### Plot a phylogeny with support values from multiple trees painted on the nodes
##### 1. `phylo_node_table()`: Create tables of node values from the phylogenies
```{r}
# Create and store ML node table
ML_node_table <- phylo_node_table(
  sauresia_ML_rooted_named,
  "ML")

# Create and store Bayesian node table 
bayesian_node_table <- phylo_node_table(
  sauresia_B_rooted_named,
  "Bayesian")
```
##### 2. `phylo_merge_nodes()`:  Merge the node tables based on matching taxa
```{r}
merged_table <- phylo_merge_nodes(
  ML_node_table,
  bayesian_node_table)
```
##### 3. `phylo_plot_tree_nodes()`: Plot a tree and paint on the merged values from `merged_table`
```{r}
phylo_plot_tree_nodes(
  tree = sauresia_ML_rooted_named,
  tree_type = "ML",
  merged_table = merged_table)
```
### 7. Documentation Functions
#### `document_get_taxonomy()`: Check the NCBI database for information on a taxon
```{r}
document_get_taxonomy("Sauresia sepsoides")
```
#### `document_material_examined()`: Create a **Materials Examined** section for each group
```{r}
document_material_examined(
  sauresia_morpho3,
  grouped_by = "Species",
  region_col = "Province",
  institution_col = "Museum",
  number_col = "Number",
  collector_col = "Collectors",
  locality_col = "Locality",
  date_collected_col = "Date Collected",
  type_col = "Type",
  sex_col = "Sex",
  lat_col = "Latitude",
  long_col = "Longitude",
  elevation_col = "Elevation")
```
#### `document_description_section()`: Create a **Description** section for each group
```{r}
document_description_section(
  sauresia_morpho3,
  trait_columns = c("SVL", "Pattern", "Head Markings", "Nuchal Lines", "Dots in Lateral Band",
                    "Mental to Vent Scale Count", "Midbody ScaleCount", "Total Finger Lamella",
                    "Strigae on Ten Scales", "Total Toe Length", "Scale Under Eye to Lip Distance",
                    "Eye Length", "Arm Length", "Ear Width", "Rostral Width", "Rostral Height",
                    "Head Length", "Mental Width", "Post Mental Width", "Cloaca Width",
                    "Prefrontal Width", "Largest Supraoccular Width", "Finger III Length",
                    "Ear to Eye Distance", "Head Width", "Frontal Width", "Frontal Length"),
  grouped_by = "Species")
```
#### `document_standout_traits()`: Identify traits that are different in one group compared to values in all or most other groups
```{r}
document_standout_traits(
  sauresia_morpho3,
  trait_columns = c("SVL", "Pattern", "Head Markings", "Nuchal Lines", "Dots in Lateral Band",
                    "Mental to Vent Scale Count", "Midbody ScaleCount", "Total Finger Lamella",
                    "Strigae on Ten Scales", "Total Toe Length", "Scale Under Eye to Lip Distance",
                    "Eye Length", "Arm Length", "Ear Width", "Rostral Width", "Rostral Height",
                    "Head Length", "Mental Width", "Post Mental Width", "Cloaca Width",
                    "Prefrontal Width", "Largest Supraoccular Width", "Finger III Length",
                    "Ear to Eye Distance", "Head Width", "Frontal Width", "Frontal Length"),
  grouped_by = "Species",
  cutoff = 2/3)
```
#### `document_diagnoses_section()`: Create a **Diagnosis** section for each group
```{r}
document_diagnoses_section(sauresia_morpho3,
                           trait_columns = c("SVL", "Pattern", "Head Markings", "Nuchal Lines", "Dots in Lateral Band",
                                             "Mental to Vent Scale Count", "Midbody ScaleCount", "Total Finger Lamella",
                                             "Strigae on Ten Scales", "Total Toe Length", "Scale Under Eye to Lip Distance",
                                             "Eye Length", "Arm Length", "Ear Width", "Rostral Width", "Rostral Height",
                                             "Head Length", "Mental Width", "Post Mental Width", "Cloaca Width",
                                             "Prefrontal Width", "Largest Supraoccular Width", "Finger III Length",
                                             "Ear to Eye Distance", "Head Width", "Frontal Width", "Frontal Length"),
                           grouped_by = "Species")
```
#### `document_trait_range_table()`: Create a table showing trait ranges for all groups
###### Including the averages and number of samples for each trait is optional
```{r}
document_trait_range_table(
  sauresia_morpho3,
  grouped_by = "Species",
  trait_columns = c("SVL", "Pattern", "Head Markings", "Nuchal Lines", "Dots in Lateral Band",
                    "Mental to Vent Scale Count", "Midbody ScaleCount", "Total Finger Lamella",
                    "Strigae on Ten Scales", "Total Toe Length", "Scale Under Eye to Lip Distance",
                    "Eye Length", "Arm Length", "Ear Width", "Rostral Width", "Rostral Height",
                    "Head Length", "Mental Width", "Post Mental Width", "Cloaca Width",
                    "Prefrontal Width", "Largest Supraoccular Width", "Finger III Length",
                    "Ear to Eye Distance", "Head Width", "Frontal Width", "Frontal Length"),
  include_counts = TRUE,
  include_averages = TRUE)
```
#### `document_appendix_section()`: Create an **Appendix** section for each group
```{r}
document_appendix_section(
  sauresia_morpho3,
  grouped_by = "Species",
  country_code_col = "Country",
  region_col = "Province",
  locality_col = "Locality",
  lat_col = "Latitude",
  long_col = "Longitude",
  elevation_col = "Elevation",
  institution_col = "Museum",
  number_col = "Number")
```

## Additional Resources

For detailed documentation on each function, see the package's help files:

```{r}
?process_column_names
?document_appendix_section
```

## Feedback and Contributions

We welcome feedback and contributions! Submit issues, suggestions, or pull requests via the GitHub repository.
